---
title: "2011 Pedestrian Volume at Toronto Intersections"
output: 
  html_document:
    keep_md: true
---

###Introduction

The following tutorial describes how to use R's [Leaflet for R](https://rstudio.github.io/leaflet/) package to develop a sharp web mapping application.  

The tutorial leverages data from Open Data Toronto - specifically, the [Signalized Intersection Traffic and Pedestrian Volume](https://www.toronto.ca/city-government/data-research-maps/open-data/open-data-catalogue/#a7b32b1a-53b7-e85b-4fb0-a4204c985a38) dataset. This dataset contains information about the quantity of pedestrians and vehicles at various intersections throughout the city in 2011, 2012, and 2013. 

* The final product of this tutorial is hosted on RPubs at: https://rpubs.com/kieg/149260

* All associated documentation can be found on Github at: https://github.com/kieg/toronto-volume 

**The final product - **By the end of the tutorial, you will have created the interactive map below. Go ahead, drag and click on the map below to investigate some of the most congested intersections in the city!

```{r, echo = FALSE, message = FALSE}

library(dplyr)   #Load our libraries
library(leaflet) #Load our libraries

files = list.files(pattern = "*.csv") #Retrieve file names

file.list <- lapply(files, 
                    function(x) read.csv(x, header = TRUE, stringsAsFactors = FALSE)) #Read files

names(file.list) <- c("vol_2011","vol_2012","vol_2013") #Name each component of list
 
#Note the invisible function keeps lapply from spitting out the data.frames to the console
invisible(lapply(names(file.list), 
                 function(x) assign(x,file.list[[x]],envir=.GlobalEnv)))

#Create intersection name and pedestrian count for an in-map pop-up.
vol_2011$Intersection <- paste(vol_2011$Main, "&", vol_2011$Side.1.Route, "|", "24hr Pedestrian Volume =", vol_2011$X24HrPedVol)


vol_2011 <- as.data.frame(lapply(vol_2011, function(x) gsub(",", "", x))) #Remove pesky commas from quantity columns


numeric_items <- vol_2011[,10:13] #Identify (soon to be) numeric columns

numeric_items <- apply(numeric_items, 2, 
                       function(x) as.numeric(x)) #Convert to numeric


vol_2011[,10:13] <- numeric_items #Put them back into our orginal dataframe

vol_2011$Longitude <- as.numeric(as.character(vol_2011$Longitude))
vol_2011$Latitude <- as.numeric(as.character(vol_2011$Latitude))

#Below I am creating a dummy variable that will be used to determine the size of the rings in the map
vol_2011$size <- ((vol_2011$X24HrPedVol - min(vol_2011$X24HrPedVol)) /
  (max(vol_2011$X24HrPedVol) - min(vol_2011$X24HrPedVol))) * 200


#Create a continuous palette function. Essentially, map our value to a series of colours to create a gradient
pal <- colorNumeric(
  palette = "Blues",
  domain = vol_2011$X24HrPedVol
)

#Other colours here: http://rstudio.github.io/leaflet/colors.html


#Below, we create the final map

map = leaflet(vol_2011) %>%   
  setView(lng = -79.381, lat = 43.656, zoom = 14) %>%        #Center the map on Toronto
  addTiles() %>%                                             #Adds in the default OpenStreetMap map tiles - we will overwrite in this case
  addProviderTiles("Hydda.Full") %>%                         #Overwrite with Hydda.Full map tile set
  addCircles(~Longitude, ~Latitude, ~size,                   #Add in circles which correspond with pedestrian traffic
                   color = ~pal(X24HrPedVol),                #Associate colour with volume
                   popup= ~Intersection,                     #Add in on-click popup functionality
                   stroke = F,                               #Remove border on circles
                   fill = TRUE,                              #Fill Circles
                   fillOpacity = .5) %>%                     #Set fill opacity
  addLegend("bottomright", pal = pal, values = ~X24HrPedVol, #Adding a legend for colour interpretation
            title = "Avg. 24hr Pedestrian Volume",
            opacity = 1
  )
```

```{r, echo = FALSE, out.width = '900px', message = FALSE}
map
```



### Step by step 

Leaflet for R makes it quite straughtforward to create visualizations like this. The steps below will walk you through each of the steps required to create this output. 

1. **Obtain data -** The first thing to do is obtain the data. The dataset is located [here](http://www1.toronto.ca/wps/portal/contentonly?vgnextoid=417aed3c99cc7310VgnVCM1000003dd60f89RCRD&vgnextchannel=1a66e03bb8d1e310VgnVCM10000071d60f89RCRD). First, convert these files to .csv format for ease and rename the 2011 file to standardize the nomenclature with the 2012 and 2013 files. 

2. **Load packages-** Next load the relevant libraries. For this particular project, only two packages are used:

**Leaflet** - the package you will be using to define and create your maps.  
**Dplyr** - a package that simplifies some syntax, allowing for the use of the pipe %>% character, among other things.


```{r, evaluate = FALSE, message = FALSE}
library(leaflet) 
library(dplyr)
```

3. **Read files:** Next, read in the relevant files. Rather than load each file individually, you can automated the process by listing the files using *list.files()*, applying the *read.csv()* command to each component of the list using *lapply()* then naming the components using *names()*. Finally, assign each component to the global environment where they can be accessed as individual dataframes.  

```{r, evaluate = FALSE, message = FALSE}
files = list.files(pattern = "*.csv") #Retrieve file names

file.list <- lapply(files, 
                    function(x) read.csv(x, header = TRUE, stringsAsFactors = FALSE)) #Read files

names(file.list) <- c("vol_2011","vol_2012","vol_2013") #Name each component of list
 
#Note the invisible function keeps lapply from spitting out the data.frames to the console
invisible(lapply(names(file.list), 
                 function(x) assign(x,file.list[[x]],envir=.GlobalEnv)))
```

4. **Pre-processing:** There are a few cleaningg steps which need to be completed prior to plotting. 

The data contains two columns to identify each intersection: **Main**, for the primary street, and **Side.1.Route**, for the secondary street. The end goal is to allow the user to identify each of this intersection, and associated pedestrian volume data via popups on the map. Use the *paste()* function to consolidate this information into a column named **Intersection**.

```{r, evaluate = FALSE}
#Create intersection name and pedestrian count for an in-map pop-up.
vol_2011$Intersection <- paste(vol_2011$Main, "&", vol_2011$Side.1.Route, "|", "24hr Pedestrian Volume =", vol_2011$X24HrPedVol)
```


Next, the average quantity of people and vehicles at each intersection will be read in as characters (rather than numeric variables) and formatted with commas. *gsub()* is used to remove the commas and *as.numeric()* to convert the numbers to the correct format (one that is usable by Leaflet).

```{r, evaluate = FALSE}
vol_2011 <- as.data.frame(lapply(vol_2011, function(x) gsub(",", "", x))) #Remove pesky commas from quantity columns

numeric_items <- vol_2011[,10:13] #Identify (soon to be) numeric columns

numeric_items <- apply(numeric_items, 2, 
                       function(x) as.numeric(x)) #Convert to numeric

vol_2011[,10:13] <- numeric_items #Put them back into our orginal dataframe

vol_2011$Longitude <- as.numeric(as.character(vol_2011$Longitude)) #Set longitude to numeric
vol_2011$Latitude <- as.numeric(as.character(vol_2011$Latitude)) #Set latitude to numeric
```
  
5. **Visualizing the information:** For this map, we will use two indicators to depict areas with higher/lower volume: 
- **circle size** (bigger = higher)  
- **colour** (darker blue = higher)  
  
To let Leaflet know how big to make each circle, use a simple equation to create a new column called **size** which will be drawn upon while generating the circles. The equation first normalizes the variable of interest, changing its range to be between 0 and 1. It then multiplies that number by 200 to enlarge the circles sizes.  

In this case, we will visualize **the average number of people over a 24 hr period**, which is named **X24HrPedVol**.

```{r, evaluate = FALSE}
# Create a dummy variable that will be used to determine the size of the rings in the map
vol_2011$size <- ((vol_2011$X24HrPedVol - min(vol_2011$X24HrPedVol)) /
  (max(vol_2011$X24HrPedVol) - min(vol_2011$X24HrPedVol))) * 200
```


To create the colour gradient, you can use the *colorNumeric()* function to map the range of values in **X24HrPedVol** to a range of colours. In this case, we will map the values to the "Blues" palette, but you can play with this yourself. For more information, you can have a look [here](http://rstudio.github.io/leaflet/colors.html)


```{r, evaluate = FALSE}
# Create a continuous palette function. Essentially, map our value to a series of colours to create a gradient
pal <- colorNumeric(
  palette = "Blues",
  domain = vol_2011$X24HrPedVol
)
```


6. **Generate the map:** That is all of the preparation that is needed! Now you are ready to use the Leafet package to create the interactive map. 

First, feed the dataframe into the leaflet function and assign it to the **map** variable. Next, modify **map** using a number of functions within the Leaflet package:

*setView()* allows you to tell Leaflet where you would initially like the map to focus, and how closely to focus on that point. 

*addTiles()* layers in the default OpenStreetMap map tiles. In this case you will use a different set, which requires the use of *addProviderTile()* instead. To see more options, you can have a look [here](https://rstudio.github.io/leaflet/basemaps.html).

*addCircleMarkers()* allows you to visualize our data by overlaying circles of varying size and colour over top of the map we have created. To alter the size or each circle based on average number of people in a 24hr period, set the radius equal to the **size** variable created earlier, then adjust the colors by setting color equal to **pal**, which contains our colour-mapped numeric values. 

*addLegend()* is fairly self-explanatory and allow you to input a legend onto your map. In this case, it specifies the quantity of people associated with circles of various colour. 

```{r, evaluate = FALSE}
#Below, we create  the final map

map = leaflet(vol_2011) %>%   
  setView(lng = -79.381, lat = 43.656, zoom = 14) %>%        #Center the map on Toronto
  addTiles() %>%                                             #Adds in the default OpenStreetMap map tiles - we will overwrite in this case
  addProviderTiles("Hydda.Full") %>%                         #Overwrite with Hydda.Full map tile set
  addCircles(~Longitude, ~Latitude, ~size,                   #Add in circles which correspond with pedestrian traffic
                   color = ~pal(X24HrPedVol),                #Associate colour with volume
                   popup= ~Intersection,                     #Add in on-click popup functionality
                   stroke = F,                               #Remove border on circles
                   fill = TRUE,                              #Fill Circles
                   fillOpacity = .5) %>%                     #Set fill opacity
  addLegend("bottomright", pal = pal, values = ~X24HrPedVol, #Adding a legend for colour interpretation
            title = "Avg. 24hr Pedestrian Volume",
            opacity = 1
  )
```



That is all! Once you call **map**, you should now see the map below.

```{r, echo = FALSE, out.width = '900px', message = FALSE}
map
```


